A modal logical multiverse for commutative algebra and combinatorics

### 2022-11-07 slides-paris2022.tex
typo
Final version

%\setbeamertemplate{headline}{\mynav{gray}{gray}{gray}}

@@

  \bigskip
  %\includegraphics[height=0.32\textwidth]{olivia-lattices}
  \bigskip
  \bigskip
  \bigskip

@@

\begin{frame}{Three take-aways}
  \bigskip
  \begin{columns}[c]
    \begin{column}{0.25\textwidth}
      \includegraphics[width=\textwidth,valign=c]{turing-machine}
    \end{column}
    \begin{column}{0.75\textwidth}
      Abstract proofs should be blueprints for concrete ones.
    \end{column}
  \end{columns}
  \pause
  \bigskip
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.25\textwidth}
      \includegraphics[width=\textwidth,valign=c]{3-adic-numbers}
    \end{column}
    \begin{column}{0.75\textwidth}
      For every inhabited set~$X$,
      there is a \hil{flavor~of~parametrized~mathematics} \\
      which is home
      to a gadget called \\
      \hil{generic surjection}~$\NN \twoheadrightarrow X$.
    \end{column}
  \end{columns}
  \pause
  \bigskip
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.25\textwidth}
      \includegraphics[width=\textwidth,valign=c]{multiverse}
    \end{column}
    \begin{column}{0.75\textwidth}
      There is a \hil{rich modal multiverse} of \\
      flavors of parametrized mathematics.
    \end{column}
  \end{columns}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
% https://tex.stackexchange.com/a/198372/32372
\newcommand\ytl[2]{%
  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
  \parbox[c]{8.4cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
  \\[-2.0pt]}
\begin{frame}{A brief timeline}
  \vspace*{-1.0em}
  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-3.5pt]
  \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
  that~$2^{\aleph_0} = \aleph_1$.}
  \pause
  \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
  \pause
  \ytl{1920s}{Set theorists pursue additional axioms to
  settle~\textsc{ch} (one way or another).}
  \pause
  \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
  is~\textsc{zfc}+\textsc{ch}.}
  \pause
  \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
  is~\textsc{zfc}+$\neg$\textsc{ch}.}
  \pause
  \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
  the philosophy of set theory.}
  \pause
  %arguing that the program of pursuing
  %additional axioms (while successful in many ways) is doomed to fail
  %to settle~\textsc{ch}.}
  \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
  mathematics.}
  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
\end{frame}
\begin{frame}{The set-theoretic multiverse}
  \begin{tikzpicture}[remember picture,overlay]
    \node[xshift=-1.5cm,yshift=-3.5cm] at (current page.north east)
    {\includegraphics[width=2cm]{branching}};
  \end{tikzpicture}
  \vspace*{-1.5em}

@@

    \smallskip
    \pause
    \justifying
    \scriptsize
    ``Let~$M$ be a finitely generated module over an
    arbitrary commutative ring~$A$. \only<1-3>{Then there is a partition~$1 = f_1 + \cdots +
    f_n \in A$ of unity such that, for each index~$i$, the localized
    module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.}\only<4>{\hil{If~$f = 0$
    is the only element of~$A$} such that~$M[f^{-1}]$ is finite free
    over~$A[f^{-1}]$, \hil{then~$1 = 0$ in~$A$}.}''
  \end{enumerate}
  \end{changemargin}
\end{frame}

@@

  \begin{changemargin}{-1.2em}{-1.2em}
    \small
    \begin{enumerate}
      %\item If a function~$\NN \to \NN$ does \emph{everywhere} \emph{not~not} have
      %a zero, then it actually has a zero.
      \item \emph{Somewhere,} the law of
      excluded middle holds. \emph{In fact, we even have:} \\
      It is \emph{everywhere} the case that the law of
      excluded middle holds \emph{somewhere}.
      \smallskip
      \item Assuming Zorn's lemma, it is
      everywhere the case that it the axiom of choice holds
      somewhere. [Barr]
      \smallskip
      \item If a geometric implication holds \emph{somewhere}, then it holds already here.
      \smallskip
      \item If a first-order statement holds \emph{proximally}, then it holds already here.
      \smallskip
      \item For every set~$X$, \emph{proximally} there is a
      surjection~$\NN \twoheadrightarrow X$.
      \smallskip
      \item Every ring \emph{proximally} has a maximal ideal.
    \end{enumerate}
  \end{changemargin}
\end{frame}}

>>>>

@@

  \bigskip
  \bigskip
  \bigskip
  \bigskip

@@

\begin{frame}{Three take-aways}
  \bigskip
  \begin{columns}[c]
    \begin{column}{0.2\textwidth}
      \includegraphics[width=\textwidth,valign=c]{turing-machine}
    \end{column}
    \begin{column}{0.75\textwidth}
      Abstract proofs should be blueprints for concrete ones.
    \end{column}
  \end{columns}
  \pause
  \bigskip
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.2\textwidth}
      \includegraphics[width=\textwidth,valign=c]{3-adic-numbers}
    \end{column}
    \begin{column}{0.75\textwidth}
      For every inhabited set~$X$,
      there is a \hil{flavor~of~parametrized~mathematics} \\
      which is home
      to a gadget called \\
      \hil{generic surjection}~$\NN \twoheadrightarrow X$.
    \end{column}
  \end{columns}
  \pause
  \bigskip
  \bigskip

  \begin{columns}[c]
    \begin{column}{0.2\textwidth}
      \includegraphics[width=\textwidth,valign=c]{multiverse}
    \end{column}
    \begin{column}{0.75\textwidth}
      There is a \hil{rich modal multiverse} of \\
      flavors of parametrized mathematics.
    \end{column}
  \end{columns}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.59cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
% https://tex.stackexchange.com/a/198372/32372
% \includegraphics[width=3cm]{lovelace-babbage}
% \includegraphics[width=3cm]{first-program}
\addtocounter{framenumber}{-1}
{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[height=\paperheight,valign=t]{ada-lovelace}\includegraphics[width=8cm,valign=t]{first-program}\end{minipage}}
\begin{frame}[plain]
  \vspace*{6cm}\hspace*{6cm}\begin{minipage}{4.7cm}
    \Large\hil{Ada Lovelace},

    \large
    the world's first \\
    computer programmer
    \medskip

    * 1815 \ \ † 1852
  \end{minipage}
\end{frame}}
\newcommand\ytl[2]{%
  \parbox[b]{4.5em}{\hfill{#1}~$\cdots\cdots$~}%
  \makebox[0pt][c]{\color{mypurple}$\bullet$}{\color{mypurple}\vrule}\quad%
  \parbox[c]{8.4cm}{\vspace{7pt}\raggedright#2\\[7pt]}%
  \\[-2.0pt]}
\begin{frame}{A brief timeline}
  \vspace*{-1.0em}
  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-3.5pt]
  \ytl{1878}{Cantor advances the \hil{continuum hypothesis}, the claim
  that~$2^{\aleph_0} = \aleph_1$.}
  \pause
  \ytl{1910s}{Zermelo--Fraenkel set theory emerges.}
  \pause
  \ytl{1920s}{Set theorists pursue additional axioms to
  settle~\textsc{ch} (one way or another).}
  \pause
  \ytl{1938}{Gödel proves: If~\textsc{zfc} is consistent, so
  is~\textsc{zfc}+\textsc{ch}.}
  \pause
  \ytl{1963}{Cohen proves: If~\textsc{zfc} is consistent, so
  is~\textsc{zfc}+$\neg$\textsc{ch}.}
  \pause
  \ytl{2011}{Hamkins offers his paper on the \hil{multiverse position} in
  the philosophy of set theory.}
  \pause
  %arguing that the program of pursuing
  %additional axioms (while successful in many ways) is doomed to fail
  %to settle~\textsc{ch}.}
  \ytl{2016}{Oldenziel proposes to study the modal multiverse of parametrized
  mathematics.}
  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
  \parbox[b]{4.5em}{\hfill\phantom{x}}\makebox[0pt][c]{\phantom{b}}{\color{mypurple}\vrule}\\[-6.5pt]
\end{frame}
\begin{frame}{The set-theoretic multiverse}
  \begin{tikzpicture}[remember picture,overlay]
    \node[xshift=-1.5cm,yshift=-3.5cm] at (current page.north east)
    {\includegraphics[width=2cm]{branching}};
  \end{tikzpicture}
  \vspace*{-1.5em}

@@

    \smallskip
    \pause
    \justifying
    \scriptsize
    ``Let~$M$ be a finitely generated module over an
    arbitrary commutative ring~$A$. \only<1-3>{Then there is a partition~$1 = f_1 + \cdots +
    f_n \in A$ of unity such that, for each index~$i$, the localized
    module~$M[f_i^{-1}]$ is finite free over~$A[f_i^{-1}]$.}\only<4>{\hil{If~$f = 0$
    is the only element of~$A$} such that~$M[f^{-1}]$ is finite free
    over~$A[f^{-1}]$, \hil{then~$1 = 0$ in~$A$}.}''\only<4>{\\\ }
  \end{enumerate}
  \end{changemargin}
\end{frame}

@@

  \begin{changemargin}{-1.2em}{-1.2em}
    \small
    \begin{enumerate}
      %\item If a function~$\NN \to \NN$ does \emph{everywhere} \emph{not~not} have
      %a zero, then it actually has a zero.
      \item \emph{Somewhere,} the law of
      excluded middle holds. \emph{In fact, we even have:} \\
      It is \emph{everywhere} the case that the law of
      excluded middle holds \emph{somewhere}.
      \smallskip
      \item Assuming Zorn's lemma, it is
      everywhere the case that it the axiom of choice holds
      somewhere. [Barr]
      \smallskip
      \item If a geometric implication holds \emph{somewhere}, then it holds already here.
      \smallskip
      \item If a first-order statement holds \emph{proximally}, then it holds already here.
      \smallskip
      \item For every inhabited set~$X$, \emph{proximally} there is a
      surjection~$\NN \twoheadrightarrow X$.
      \smallskip
      \item Every ring \emph{proximally} has a maximal ideal.
    \end{enumerate}
  \end{changemargin}
\end{frame}}

\backupstart

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\includegraphics[width=\paperwidth]{eventually-monad}\end{minipage}}
\begin{frame}
  \vspace*{\paperheight}\vspace*{-3.4em}\centering
  \emph{Agda formalization available.}
\end{frame}}

\backupend
